---
title: "Assignment 5: Data Visualization"
author: "Asreeta Ushasri"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Install packages and read in data

#install.packages(tidyverse)
#install.packages(cowplot)
#install.packages(lubridate)
#install.packages(here)
#install.packages(ggridges)

library(ggridges)
library(tidyverse)
library(lubridate)
library(here)

getwd()

here()

PeterPaul <- read.csv(
  file = here("Data/Processed/NTL_LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

Litter <- read.csv(
  file = here("Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE)

#view(PeterPaul)
#view(Litter)

#2 Check the date format
class(PeterPaul$sampledate)
class(Litter$sampledate)

# Format sampledate as date
PeterPaul$sampledate <- as.Date(PeterPaul$sampledate, format = "%Y-%m-%d")

Litter$sampledate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3 Creating a plot theme

#install.packages(viridisLite)

library(viridis)
library(RColorBrewer)
library(colormap)
library(ggthemes)


#Create a custom theme, using elements from lab document

my_theme <- theme_tufte() +
theme(
  line = element_line(),
  rect = element_rect(),
  text = element_text(),
  

#Text Element
  plot.title = element_text("black"),
  axis.title.x = element_text("black"),
  axis.title.y = element_text("black"),
  axis.text = element_text("black"),

#Line Element
  axis.ticks = element_line("gray"),
  panel.grid.major = element_line("lightgray"),
  panel.grid.minor = element_line("lightgray"),

#Rectangle Element
  plot.background = element_rect("white"),
  panel.background = element_rect("white"),
  legend.key = element_rect("white"),

#Legend Position
  legend.position = 'right',
complete = TRUE)


#Test My Default theme
PeterPaul %>% 
  ggplot() + 
  geom_bar(aes(x=month)) + my_theme

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 Plot total phosphorus by phosphate in Peter and Paul Lakes.

#First, set up basic line plot
PeterPaulPlot <- PeterPaul %>%
  ggplot(mapping =
         aes(x=po4,
             y=tp_ug,
             color=lakename)) + geom_line()

#show(PeterPaulPlot)

#Change scale of x and y axis
PeterPaulPlotLimits <- PeterPaulPlot + xlim(NA, 45) + ylim(NA, 150)

#show(PeterPaulPlotLimits)

#Add axis titles
PeterPaulPlotTitles <- PeterPaulPlotLimits +
  labs(
    title = "Total Phosphorus by Phosphate
    in North Temperate Wisconsin Lakes",
    x = "Phosphate (PO4)",
    y = "Total Phosphorus",
    color = "Lake Name"
    ) + theme(axis.title.y = element_text(angle = 90, vjust = 0.5, hjust=1)) 

#show(PeterPaulPlotTitles)


#Add line of best fit and custom theme

Peter_Paul_Phos_Graph <- PeterPaulPlotTitles +
  geom_smooth(method = lm,
              se=FALSE) + my_theme +
  theme(axis.title.y = element_text(angle = 90))

show(Peter_Paul_Phos_Graph)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 Create Boxplots

#First, create temperature boxplot

Peter_Paul_Temp <- ggplot(data = PeterPaul,
       aes(
         x=month,
         y=temperature_C,
         color=lakename)) +
  geom_boxplot() +
  labs(
    title = "Temperature
    in Celsius",
    x = element_blank(),
    y = element_blank()) +
  theme(legend.position = "none")

show(Peter_Paul_Temp)

#Next, create phosphorus boxplot

Peter_Paul_TP <- ggplot(data = PeterPaul,
       aes(
         x=month,
         y=tp_ug,
         color=lakename)) +
  geom_boxplot() +
  labs(
    title = "Total
    Phosphorus",
    x = element_blank(),
    y = element_blank()) +
  theme(legend.position = "none")

show(Peter_Paul_TP)

#Lastly, create nitrogen boxplot

Peter_Paul_TN <- ggplot(data = PeterPaul,
       aes(
         x=month,
         y=tn_ug,
         color=lakename)) +
  geom_boxplot() +
  labs(
    title = "Total
    Nitrogen",
    x = element_blank(),
    y = element_blank(),
    color = "Lake Name") + 
  theme(legend.position = "right")


show(Peter_Paul_TN)

#Combine multiple plots

library(cowplot)

Boxplots <-
  plot_grid(Peter_Paul_Temp,
            Peter_Paul_TP, Peter_Paul_TN,
            ncol = 3,
            align = 'h', axis = 'tblr',
            rel_widths = c(1, 1, 1.5))

show(Boxplots)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Peter Lake has slightly lower temperatures than Paul Lake, shown by a lower minimum, median, and maximum value. In contrast, Peter Lake has higher median phosphorus and nitrogen levels. The phosophorus and nitrogen graphs appear to contain a variety of outliers at the top of the boxplots. This indicates that the distribution of phosphorus and nitrogen for both Peter and Paul Lakes are skewed right.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6 Plotting Needles Group Dry Mass

#First, create new dataset filtered by Needles
Litter_Needles <- Litter %>% 
  filter(functionalGroup == "Needles")

#Now, create plot from Litter_Needles
ggplot(data = Litter_Needles,
       aes(x = collectDate, y = dryMass,
           color = nlcdClass)) + geom_point() +
  labs(
    title = "Dry Mass of Needles Litterfall at NEON Sites",
    x = "Sample Collection Date",
    y = "Dry Mass",
    color = "NLCD Class") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#7 Changing NLCD Class from color to facet

#Use the Litter_Needles dataset
ggplot(data = Litter_Needles,
       aes(x = collectDate, y = dryMass,
           facet_grid = nlcdClass)) + geom_point() +
  labs(
    title = "Dry Mass of Needles Litterfall at NEON Sites",
    x = "Sample Collection Date",
    y = "Dry Mass",
    facet_grid = "NLCD Class") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 6 is more effective than Plot 7, because Plot 6 displays the different NLCD classes on the same graph. This allows researchers to compare and contrast the dry mass values of evergreen forest samples, grassland herbaceous samples, and shrub scrub samples easily with color differentiation. In Plot 6, we can see that the evergreen forest and grassland herbaceous samples had higher dry mass values from October 2016 to October 2017. The Shurb Scrub dry mass remained lower than the other two NLCD groups throughout this time frame. Plot 7 does not include any color differentiation, which makes it difficult to determine which scatterplot points represent the three different NLCD classes.